---
title: "How to easily split transects into segemnts using the dshm-package"
author: "Filippo Franchini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
#save(raw.trans,file = "/Users/User/Desktop/dshm/datasets/data_split_transects.Rdata")
load("/Users/User/Desktop/dshm/datasets/data_split_transects.RData")
```

In this tutorial you will learn how to use the `dshm` R-package to:

- Split transects into segments and apply a buffer to them using the `dshm_split_transects` function
- Check if segments have overlapping areas using the `dshm_check_segments` function. You can decide to correct the segments using the function `dshm_correct_segments` and re-check to proof all the overlapping areas have been deleted.
- Finalize the segments for spatial analyisis using the function `dshm_finalize_segments`. This process includes:

    - Taking out overlapping areas between buffered segments and land
    - Calculating covariate statistics for each segment


## Splitting transects into segments

The object `raw.trans` is a `SptialLinesDataFrame` containign 72 lines (i.e. transects) with data associated with each line that you can access by typing:
      
    raw.trans@data

You can show the `raw.trans` data by typing

    raster::plot(raw.trans)

This should produce a plot like this one

```{r}
raster::plot(raw.trans)
```

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
