---
title: "How to easily split transects into segemnts using the dshm-package"
author: "Filippo Franchini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=300
)
#save(transects,file = "/Users/User/Desktop/dshm/datasets/data_split_transects.Rdata")
load("/Users/User/Desktop/dshm/datasets/data_split_transects.RData")
```

In this tutorial you will learn how to use the `dshm` R-package to:

- Split transects into segments and apply a buffer to them using the `dshm_split_transects` function
- Check if segments have overlapping areas using the `dshm_check_segments` function. You can decide to correct the segments using the function `dshm_correct_segments` and re-check to proof all the overlapping areas have been deleted.
- Finalize the segments for spatial analyisis using the function `dshm_finalize_segments`. This process includes:

    - Taking out overlapping areas between buffered segments and land
    - Calculating covariate statistics for each segment


## Splitting transects into segments

The object `raw.trans` is a `SptialLinesDataFrame` containign 72 lines (i.e. transects) with data associated with each line that you can access by typing:
      
    transects@data

You can show the `raw.trans` data by typing

```{r, echo=FALSE, out.width = "300px"}
raster::plot(transects)
```
